Really Hard Instances for the Subgraph Isomorphism Problem

### 2015-11-14 paper/paper.tex
More

\usepackage{ijcai13}
\usepackage{times}
\usepackage{complexity}
\usepackage{microtype}
\usepackage{gnuplot-lua-tikz}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{placeins}
\usepackage{cleveref}
\usepackage{natbib}

% \usepackage{showframe}
\usepackage{lipsum}

@@

\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}

@@

However, generating non-trivial unsatisfiable random instances is not straightforward. Using a
pattern graph from one of the random suites with the ``wrong'' target graph tends to give either a
trivially unsatisfiable instance, or a satisfiable instance. (In particular, it is \emph{not} the
case that a relatively small random graph is unlikely to appear in a larger random graph.)

\begin{figure}[b]
    \centering
    \begin{tikzpicture}[scale=0.5]
        \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (Na) at (1,  0) {1};
        \node [draw, circle, fill=uofgpumpkin, inner sep=1.5pt] (Nb) at (1, -1.5) {2};
        \node [draw, circle, fill=uofglawn, inner sep=1.5pt] (Nc) at (0, -3) {3};
        \node [draw, circle, fill=uofgthistle, inner sep=1.5pt] (Nd) at (2, -3) {4};

@@

        \node [anchor=center, font=\Large] (A1) at (-1.5, -1.5) { $\hookleftarrow$ };
        \node [anchor=center, font=\Large] (A2) at ( 3.5, -1.5) { $\rightarrowtail$ };
    \end{tikzpicture}

    \caption{There is a non-induced isomorphism from the pattern graph, in the center, to the target
    on the right, mapping vertex 1 to 1, 2 to 3, 3 to 5 and 4 to 6. This is not an induced
    isomorphism, since there is an edge between 1 and 5 in the target but not between 1 and 3 in the
    pattern. The mapping from the pattern to the (same) target on the left, sending 1 to 2, 2 to 6, 3 to
    5 and 4 to 3, is both non-induced and induced.}
    \label{figure:sip}
\end{figure}

\begin{figure}[t]
    \input{gen-graph-phase-transition.tex}
    \caption{With a fixed a pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.4, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance. The lines show mean search effort, and mean proportion satisfiable, using a
    larger sample size.}
    \label{figure:phase-transition}
\end{figure}

To predict the location of the phase transition, we can calculate the expected number of solutions
by making a few dubious assumptions involving rounding and independence. Since we are trying to find
an \emph{injective} mapping from a pattern $P = G(p, d_p)$ to a target $T = G(t, d_t)$, there are
$t^{\underline{p}} = t \cdot (t - 1)\cdots(t - p + 1)$ possible assignments of target vertices to
pattern vertices.  We expect the pattern to have $d_p \cdot \binom{p}{2}$ edges, so we obtain the
probability of all of these edges being mapped to edges in the target by raising $d_t$ to this
power, giving an expected number of solutions of \[ t^{\underline{p}} \cdot {d_t}^{d_p \cdot
\binom{p}{2}} \textnormal{,} \] which we plot in ?? somewhere.

\begin{figure*}[h]
    \input{gen-graph-induced.tex}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The second row shows a bound on the satisfiable region, by
    considering where a \emph{non-}induced isomorphism may also be a non-induced isomorphism between
    complement graphs.}\label{figure:induced}
\end{figure*}

By repeating the argument for locating the non-induced phase transition and additionally considering
non-edges, we get an expected number of solutions of \[t^{\underline{p}} \cdot {d_t}^{d_p \cdot
\binom{p}{2}} \cdot {d_{\overline{t}}}^{d_{\overline{p}} \cdot \binom{p}{2}} \textnormal{,} \]
where $d_{\overline{p}} = 1 - d_p$ and $d_{\overline{t}} = 1 - d_t$. We plot this ?? somewhere.

\begin{figure}[tb]
    \centering
    \begin{tikzpicture}[scale=0.5]

        \foreach \n in {1, ..., 6}{
            \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (N1\n) at (\n, 0) {\n};
        }

@@

            \draw [thick] ($(N12)!0.8!(N12.south)$) -- ($(N43)!0.8!(N43.north)$);
            \draw [thick] ($(N12)!0.8!(N12.south)$) -- ($(N35)!0.8!(N35.west)$);
            \draw [thick] ($(N12)!0.8!(N12.south)$) -- ($(N26)!0.8!(N26.east)$);
            \draw [thick] ($(N43)!0.8!(N43.north)$) -- ($(N35)!0.8!(N35.west)$);
            \draw [thick] ($(N43)!0.8!(N43.north)$) -- ($(N26)!0.8!(N26.east)$);
            \draw [thick] ($(N35)!0.8!(N35.west)$) -- ($(N26)!0.8!(N26.east)$);

@@

    \end{tikzpicture}
    \caption{The association graph encoding of the induced version of \cref{figure:sip}. The clique
    of size four corresponds to the mapping sending pattern vertex 1 to target vertex 2, 2 to 6, 3
    to 5, and 4 to 3.}\label{figure:assoc}
\end{figure}

>>>>

\usepackage{ijcai13}
\usepackage{times}
\usepackage{complexity}
\usepackage{microtype}
\usepackage{gnuplot-lua-tikz}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{placeins}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{dblfloatfix}

% \usepackage{showframe}

@@

\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}
\renewcommand{\dbltopfraction}{0.9}
\renewcommand{\textfraction}{0.05}
\setcounter{dbltopnumber}{2}
\renewcommand{\floatpagefraction}{0.85}
\renewcommand{\dblfloatpagefraction}{0.85}

@@

\begin{figure}[b]
    \centering
    \begin{tikzpicture}[scale=0.5]
        \begin{scope}
            \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (Na) at (1,  0) {1};
            \node [draw, circle, fill=uofgpumpkin, inner sep=1.5pt] (Nb) at (1, -1.5) {2};
            \node [draw, circle, fill=uofglawn, inner sep=1.5pt] (Nc) at (0, -3) {3};
            \node [draw, circle, fill=uofgthistle, inner sep=1.5pt] (Nd) at (2, -3) {4};

@@

            \node [anchor=center, font=\Large] (A1) at (-1.5, -1.5) { $\hookleftarrow$ };
            \node [anchor=center, font=\Large] (A2) at ( 3.5, -1.5) { $\rightarrowtail$ };
        \end{scope}
    \end{tikzpicture}\\[0.5cm]
    \begin{tikzpicture}[scale=0.5]
        \begin{scope}
            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (N1\n) at (\n, 0) {\n};
            }

@@

                \draw [ultra thick] ($(N12)!0.8!(N12.south)$) -- ($(N43)!0.8!(N43.north)$);
                \draw [ultra thick] ($(N12)!0.8!(N12.south)$) -- ($(N35)!0.8!(N35.west)$);
                \draw [ultra thick] ($(N12)!0.8!(N12.south)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [ultra thick] ($(N43)!0.8!(N43.north)$) -- ($(N35)!0.8!(N35.west)$);
                \draw [ultra thick] ($(N43)!0.8!(N43.north)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [ultra thick] ($(N35)!0.8!(N35.west)$) -- ($(N26)!0.8!(N26.east)$);

@@

    \end{tikzpicture}

    \caption{On top, there is a non-induced isomorphism from the pattern graph, in the center, to the target
    on the right, mapping vertex 1 to 1, 2 to 3, 3 to 5 and 4 to 6. This is not an induced
    isomorphism, since there is an edge between 1 and 5 in the target but not between 1 and 3 in the
    pattern. The mapping from the pattern to the (same) target on the left, sending 1 to 2, 2 to 6, 3 to
    5 and 4 to 3, is both non-induced and induced.
    Below, the association graph encoding for the induced version. The bold clique of size
    four corresponds to the same mapping.}
    \label{figure:sip}
\end{figure}

Generating non-trivial unsatisfiable random instances is not straightforward. Using a
pattern graph from one of the random suites with the ``wrong'' target graph tends to give either a
trivially unsatisfiable instance, or a satisfiable instance. (In particular, it is \emph{not} the
case that a relatively small random graph is unlikely to appear in a larger random graph.)

The \emph{association graph encoding} of a subgraph isomorphism problem is constructed by creating a
new graph with a vertex for each pair $(p, t)$ of vertices from the pattern and target graphs
respectively. There is an edge between vertex $(p_1, t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$
to $t_1$ and $p_2$ to $t_2$ simultaneously is permitted. For the non-induced problem, permitted
means either $p_1$ is not adjacent to $p_2$, or both $p_1$ and $p_2$, and $t_1$ and $t_2$, are
adjacent; for the induced problem, it means that $p_1$ is adjacent to $p_2$ if and only if $t_1$ is
adjacent to $t_2$. Either way, it can be seen that a clique (a subset of vertices, each of which is
adjacent to every other in the subset) of size equal to the order of the pattern graph exists in the
association graph if and only if the problem is satisfiable (and the vertices selected correspond to
the assignments of the solution).

?? Cite max common subgraph.

\begin{figure}[tb]
    \input{gen-graph-phase-transition.tex}
    \caption{With a fixed a pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.4, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance. The lines show mean search effort, and mean proportion satisfiable, using a
    larger sample size.}
    \label{figure:phase-transition}
\end{figure}

To predict the location of the phase transition, we can calculate the expected number of solutions
by making a few dubious assumptions involving rounding and independence, and drawing inspiration
from the association graph encoding. Since we are trying to find an \emph{injective} mapping from a
pattern $P = G(p, d_p)$ to a target $T = G(t, d_t)$, there are $t^{\underline{p}} = t \cdot (t -
1)\cdots(t - p + 1)$ possible assignments of target vertices to pattern vertices.  We expect the
pattern to have $d_p \cdot \binom{p}{2}$ edges, so we obtain the probability of all of these edges
being mapped to edges in the target by raising $d_t$ to this power, giving an expected number of
solutions of \[ \langle Sol \rangle = t^{\underline{p}} \cdot {d_t}^{d_p \cdot \binom{p}{2}}
\textnormal{,} \] which we plot in ?? somewhere.

\begin{figure*}[tb]
    \input{gen-graph-induced.tex}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The second row shows a bound on the satisfiable region, by
    considering where a \emph{non-}induced isomorphism may also be a non-induced isomorphism between
    complement graphs.}\label{figure:induced}
\end{figure*}

By repeating the argument for locating the non-induced phase transition and additionally considering
non-edges, we get an expected number of solutions of \[ \langle Sol \rangle = t^{\underline{p}}
\cdot {d_t}^{d_p \cdot \binom{p}{2}} \cdot {d_{\overline{t}}}^{d_{\overline{p}} \cdot \binom{p}{2}}
\textnormal{,} \] where $d_{\overline{p}} = 1 - d_p$ and $d_{\overline{t}} = 1 - d_t$. We plot this
?? somewhere.
